---
title: "Getting started with PGEcore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with PGEcore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

PGEcore is an R package for malaria genomics analysis. It **wraps existing
tools** so they share a consistent R API, CLI, and TSV formats, and it also
**adds extra pieces**—for example naive COI, frequency, and prevalence methods,
filters, and converters—that sit alongside those wrappers.

Every tool is available both as an **R function** and as a **command-line
program**, so you can use it interactively, in scripts, or as a standard
library across workflow pipelines. Shared table layouts make it easier to move
results from one analysis into the next without reformatting.

## Install

```{r}
# remotes::install_github("PlasmoGenEpi/PGEcore")
library(PGEcore)
```

Specialised packages used by some wrappers (for example `dcifer`, `moire`) are
**Suggests** and must be installed separately when you need those tools.

## From R

```{r}
library(PGEcore)

# File paths
count_samples_by_coi("my_coi_table.tsv", output = "coi_distribution.tsv")

# Or in-memory tables
coi_table <- readr::read_tsv("my_coi_table.tsv", show_col_types = FALSE)
count_samples_by_coi(coi_table)
```

Function names match the CLI name (`count_samples_by_coi`, `moire_wrapper`,
…). That is the API to start with. A small number of wrappers also export an
in-memory `run_*` helper (`run_coiaf`, `run_moire`, `run_malariaem`) for when
you already have data frames or model objects in R; see `?run_moire`.

```{r}
?count_samples_by_coi
help(package = "PGEcore")
```

## Command line

Each tool has an executable in the package `exec/` directory. After an install
that puts those on `PATH`:

```bash
count_samples_by_coi \
  --coi_table my_coi_table.tsv \
  --output coi_distribution.tsv
```

From a source checkout of this repository:

```bash
Rscript exec/count_samples_by_coi \
  --coi_table inst/extdata/example_coi_table.tsv \
  --output coi_distribution.tsv

Rscript exec/estimate_coi_naive \
  --allele_table inst/extdata/example_allele_table.tsv \
  --output coi_table.tsv

Rscript exec/dcifer_slaf_wrapper \
  --allele_table inst/extdata/example_allele_table.tsv \
  --slaf_output slaf.tsv
```

Pass ordinary file paths. Example TSVs that show the expected columns are under
`inst/extdata/` (see the **input-formats** vignette).

## Combining analyses

1. Match your tables to a **standard format** (SNP calls, allele table, amino
   acid calls, COI table, …).
2. Run a PGEcore function or CLI.
3. Feed the output TSV into the next tool when the column layouts already align.

That interoperability—one invocation style and shared table layouts—is a core
goal of the package.
