---
title: "Standard input and output formats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard input and output formats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

PGEcore tools read and write **TSV** tables with agreed column names so outputs
from one step can be inputs to another. Function and CLI arguments use the same
schema names as the sections below (`allele_table`, `aa_calls`, `snp_calls`,
`coi_table`, `loci_groups`, …). Tool help pages describe *which* tables they
need; this vignette is the place for column layouts and examples.

Bundled examples live in `inst/extdata/` after you install the package (or in
that folder in the source tree).

```{r}
library(PGEcore)
extdata <- system.file("extdata", package = "PGEcore")
list.files(extdata, pattern = "^example")[1:8]
```

The snippets below use those examples only to show **column structure**. In
normal use, pass paths to **your** files on the CLI or in R.

## COI table (`coi_table`)

**Columns:** `specimen_name`, `coi`

**Example:** `example_coi_table.tsv`  
**Used by:** `count_samples_by_coi`, and as optional COI input to several
wrappers (`dcifer_*`, …).  
**Exception:** `FreqEstimationModel_wrapper` uses `--coi`, which accepts either
a path to this table or a single numeric average COI.

```{r}
readr::read_tsv(
  file.path(extdata, "example_coi_table.tsv"),
  show_col_types = FALSE,
  n_max = 3
)
```

```bash
Rscript exec/count_samples_by_coi \
  --coi_table my_coi_table.tsv \
  --output coi_distribution.tsv
```

## SNP calls (`snp_calls`)

**Minimum columns:** `specimen_name`, `snp_name`, `reads`, `seq_base`

Collapsed SNP tables may include extra columns (`target_name`, `chrom`, `pos`,
…). Tools that need only the minimum set ignore the rest.

**Example:** `example_collapsed_snp_calls.tsv`  
**Used by:** `coiaf_wrapper`, `THEREALMcCOIL_wrapper`, `snp_calls_to_vcf`,
SNP filtering and pileup tools, …

```{r}
readr::read_tsv(
  file.path(extdata, "example_collapsed_snp_calls.tsv"),
  show_col_types = FALSE,
  n_max = 3
) |>
  dplyr::select(specimen_name, snp_name, reads, seq_base)
```

```bash
Rscript exec/THEREALMcCOIL_wrapper \
  --snp_calls my_snp_calls.tsv \
  --slaf_output slaf.tsv \
  --coi_output coi.tsv
```

## Allele table (`allele_table`)

**Columns:** `specimen_name`, `target_name`, `seq`, `reads`

**Example:** `example_allele_table.tsv`  
**Used by:** `estimate_coi_naive`, `dcifer_slaf_wrapper`, `moire_wrapper`,
`malariaem_wrapper`, `allele_per_locus_summary`, `per_locus_popgen_summary`, …

```{r}
readr::read_tsv(
  file.path(extdata, "example_allele_table.tsv"),
  show_col_types = FALSE,
  n_max = 3
)
```

```bash
Rscript exec/dcifer_slaf_wrapper \
  --allele_table my_allele_table.tsv \
  --slaf_output slaf.tsv
```

## AA calls (`aa_calls`)

**Core columns:** `specimen_name`, `gene_id`, `aa_position`, `aa`, `reads`  
Often also: `target_name`, `aa_locus`, `gene`, `ref_aa`

**Example:** `example_aa_calls.tsv`  
**Used by:** `estimate_allele_frequency_naive`, `estimate_allele_prevalence_naive`,
`filter_biallelic_calls`, `multilocus_prevfreq_*`, `snpslice_wrapper`, …

```{r}
readr::read_tsv(
  file.path(extdata, "example_aa_calls.tsv"),
  show_col_types = FALSE,
  n_max = 3
)
```

## Loci groups (`loci_groups`)

**Columns:** `group_id`, `gene_id`, `aa_position` (often `aa_locus` too)

Defines which loci are analysed together for multi-locus tools.

**Example:** `example_loci_groups.tsv`  
**Used by:** `multilocus_prevfreq_*`, `snpslice_wrapper`,
`FreqEstimationModel_wrapper`, `MultiLociBiallelicModel_wrapper`

```{r}
readr::read_tsv(
  file.path(extdata, "example_loci_groups.tsv"),
  show_col_types = FALSE,
  n_max = 4
)
```

## Multilocus allele frequency (MLAF)

**Columns:** `group_id`, `variant`, `freq`

`variant` is often a STAVE-style string (`gene_id:aa_position:aa`).

**Example:** `example_mlaf.tsv`  
**Used by:** `slaf_from_stave_mlaf` (and produced by several multi-locus tools)

```{r}
readr::read_tsv(
  file.path(extdata, "example_mlaf.tsv"),
  show_col_types = FALSE,
  n_max = 3
)
```

## Population-level minor allele frequency (PLMAF)

**Columns:** `snp_name`, `seq_base`, `plmaf`

**Example:** `example_coiaf_plmaf.tsv`  
**Used by:** `coiaf_wrapper` (optional; otherwise PLMAF is estimated from the
SNP table)

```{r}
readr::read_tsv(
  file.path(extdata, "example_coiaf_plmaf.tsv"),
  show_col_types = FALSE,
  n_max = 3
)
```

## Microhaplotype SLAF

**Columns:** `target_name`, `seq`, `freq`, `sample_total`

**Example:** `example_mhaps_slaf.tsv`  
**Used by:** `slaf_from_mhaps_freqs`

```{r}
readr::read_tsv(
  file.path(extdata, "example_mhaps_slaf.tsv"),
  show_col_types = FALSE,
  n_max = 3
)
```

## Common outputs

| Output | Typical columns | Example producers |
| ------ | --------------- | ----------------- |
| COI per specimen | `specimen_name`, `coi` | `estimate_coi_naive`, McCOIL, snpslice, … |
| COI distribution | `coi`, `n`, `proportion` | `count_samples_by_coi` |
| Allele frequency | `variant`, `freq` (sometimes counts) | naive AF, McCOIL SLAF, … |
| Prevalence | `variant`, `prev`, … | `estimate_allele_prevalence_naive` |

When you add a new tool, reuse these layouts whenever possible so downstream
steps do not need format converters.
